---
title: "Video Game Film Adaptations"
author: "Harsh, Charlie, Mason"
date: "2026-06-15"
format: pdf
---

```{r setup}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(readr)
library(janitor)
library(scales)

ggplot2::theme_set(theme_minimal(base_size = 12))
```

## Introduction

[Intro goes here]

## Methodology

### Q1: Critic–Audience Agreement Over Time

Do critics and audiences agree on video game movies, and has that gap changed over time?

Compare `rotten_tomatoes` and `metacritic` against `cinema_score` (convert letter grades to numeric scale, drop `"N/A"` rows). Also plot the critic/audience gap across `release_date` to see if it narrows over time.

### Q2: Publisher Size & Box Office

Does publisher size predict box office success?

Join external publisher revenue/size data on `original_game_publisher`. Filter to USD only (`worldwide_box_office_currency == "$"`). Adjust `worldwide_box_office` for inflation to 2024 dollars via a CPI table joined on year extracted from `release_date`.

### Q3: Franchise Fatigue

Does franchise fatigue exist in video game movie series?

Use regex on `title` to detect both numbered sequels (`"Resident Evil 2"`) and subtitle-based sequels (`"Resident Evil: Apocalypse"`). Assign installment numbers within each franchise sorted by `release_date`. Focus on franchises with 3+ films. Plot inflation-adjusted USD box office and `rotten_tomatoes` by installment number.

### Known Data Issues

- Heavy NAs across score columns (`rotten_tomatoes`, `metacritic`, `cinema_score`)
- Mixed currencies in `worldwide_box_office` and `budget_low`/`budget_high`
- No inflation adjustment in raw data
- `cinema_score` is a character column needing numeric conversion before analysis

## Data Import

```{r import}
#| label: import

game_films <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-06-09/game_films.csv",
  show_col_types = FALSE
) |>
  janitor::clean_names()

glimpse(game_films)
```

### Data Preparation: Currency Standardization

All monetary columns are in their *original* currency. We standardize to USD.

```{r currency-lookup}
#| label: currency-lookup

fx_rates <- tibble::tribble(
  ~currency, ~to_usd,
  "$",        1,
  "¥",        0.0091,
  "£",        1.30,
  "HK$",      0.128,
  "NT$",      0.033
)
```

```{r currency-convert}
#| label: currency-convert

game_films <- game_films |>
  left_join(fx_rates, by = c("worldwide_box_office_currency" = "currency")) |>
  mutate(
    worldwide_box_office_usd = worldwide_box_office * to_usd
  ) |>
  select(-to_usd) |>
  left_join(fx_rates, by = c("budget_currency" = "currency")) |>
  mutate(
    budget_low_usd  = budget_low  * to_usd,
    budget_high_usd = budget_high * to_usd
  ) |>
  select(-to_usd)

summary(game_films$worldwide_box_office_usd)
```

### Data Preparation: Inflation Adjustment

Adjust all USD monetary columns to constant 2024 dollars using CPI-U.

```{r cpi-table}
#| label: cpi-table

cpi <- tibble::tribble(
  ~year, ~cpi,
  1986, 109.6,
  1987, 113.6,
  1988, 118.3,
  1989, 124.0,
  1990, 130.7,
  1991, 136.2,
  1992, 140.3,
  1993, 144.5,
  1994, 148.2,
  1995, 152.4,
  1996, 156.9,
  1997, 160.5,
  1998, 163.0,
  1999, 166.6,
  2000, 172.2,
  2001, 177.1,
  2002, 179.9,
  2003, 184.0,
  2004, 188.9,
  2005, 195.3,
  2006, 201.6,
  2007, 207.3,
  2008, 215.3,
  2009, 214.5,
  2010, 218.1,
  2011, 224.9,
  2012, 229.6,
  2013, 233.0,
  2014, 236.7,
  2015, 237.0,
  2016, 240.0,
  2017, 245.1,
  2018, 251.1,
  2019, 255.7,
  2020, 258.8,
  2021, 271.0,
  2022, 292.7,
  2023, 304.7,
  2024, 313.7
)

cpi_2024 <- cpi$cpi[cpi$year == 2024]
```

```{r inflation-adjust}
#| label: inflation-adjust

game_films <- game_films |>
  mutate(release_year = lubridate::year(release_date)) |>
  left_join(cpi, by = c("release_year" = "year")) |>
  mutate(
    inflation_factor = cpi_2024 / cpi,
    worldwide_box_office_usd_adj = worldwide_box_office_usd * inflation_factor,
    budget_low_usd_adj  = budget_low_usd  * inflation_factor,
    budget_high_usd_adj = budget_high_usd * inflation_factor
  ) |>
  select(-cpi, -inflation_factor)

game_films |>
  filter(!is.na(worldwide_box_office_usd)) |>
  summarise(
    n                    = n(),
    raw_median           = median(worldwide_box_office, na.rm = TRUE),
    usd_median           = median(worldwide_box_office_usd, na.rm = TRUE),
    adj_median           = median(worldwide_box_office_usd_adj, na.rm = TRUE)
  )
```

### Data Preparation: Movies Only!

Keep only entries that are theatrical releases. All others lack crucial information about budget, gross earnings, and review scores.

```{r theatrical-release}
#| label: theatrical-release

game_films <- game_films %>% 
  filter(category == "Theatrical release")

```

### Data Preparation: Cinemascore conversion

converts the Cinemascore variable to a numeric score

```{r cinema-convert}
#| label: cinema-convert

cinema_score_conversion <- tibble::tribble(
  ~cinemascore_letter, ~to_num,
  "N/A",        NA,
  "A+",        100,
  "A",        91.6666,
  "A-",      83.3333,
  "B+",      74.9999,
  "B",        66.6666,
  "B-",        58.3333,
  "C+",      49.9999,
  "C",      41.6666,
  "C-",        33.3333,
  "D+",      24.9999,
  "D",      16.6666,
  "D-",        8.3333,
  "F",        0,
)

game_films <- game_films %>% 
  left_join(cinema_score_conversion, join_by("cinema_score" == "cinemascore_letter"))

game_films <- game_films %>%
  mutate(cinema_score = to_num) %>% 
  select(-starts_with("to_num"))

```

### Cleaned Dataset Summary

```{r final-glimpse}
#| label: final-glimpse

glimpse(game_films)

game_films |>
  summarise(
    total_films        = n(),
    n_with_box_office  = sum(!is.na(worldwide_box_office)),
    n_with_rt          = sum(!is.na(rotten_tomatoes)),
    n_with_cinema      = sum(!is.na(cinema_score) & cinema_score != "N/A"),
    date_range          = paste(range(release_date, na.rm = TRUE), collapse = " to ")
  )
```

## Question 1: Critic vs. Audience Review Scores Over Time

[Reintroduce question]

[Insert Data Table]

[Explain Data Table]

[Insert Plot]

[Explain Plot]

[Concluding Statements]

## Question 2: Publisher Size vs. Box Office

  Aside from review scores, another question that interested us was determining if the size of a game's company was a predictor of the movie's success. Do dominant companies like Nintendo and Sony triumph over more niche businesses such as Capcom and Ubisoft? Let's find out.

  In order to wrangle the data into a form suitable for answering this question, we need to find data that actually represents the size of a company. It would be unfeasible to manually create a supplement data frame containing every company included in game_films, so instead we look to the wider internet!

  If we reasonably assume that larger companies are more likely to spend the money to create movies of their IPs, then perhaps a list of the fifty most successful video game companies will share an adequate amount of names with our data. That is why we turned to Wikipedia's list of that metric. It can be found here:
  
> https://en.wikipedia.org/wiki/List_of_largest_video_game_companies_by_revenue

  After scraping the data and joining it to game_films, we were left with 36 matched entries after filtering out all entries with missing data. A list of this table can be seen below.

```{r company-size-table}
company_url = "https://en.wikipedia.org/wiki/List_of_largest_video_game_companies_by_revenue"
TablesRedun <- company_url |>
  read_html() |>
  html_elements("table") |>
  html_table()
top_50 <- TablesRedun[[2]]
top_50 <- clean_names(top_50) %>% 
  mutate(revenue_usd_billions = parse_number(revenue_usd)) %>% 
  select(!c(ref, revenue_usd))
top_50 <- as.data.frame(top_50) %>% 
  right_join(game_films, join_by("company" == "original_game_publisher"), copy) %>% 
  select(title, worldwide_box_office, company, revenue_usd_billions) %>% 
  filter(!is.na(worldwide_box_office), !is.na(revenue_usd_billions)) %>% 
  arrange(title)
top_50 %>% 
  head(10)
```

  Taking a look at the first ten entries of this table, we can see that the resulting data frame includes companies ranging from Ubisoft to Sega to Square Enix and beyond. In order to get a clear picture of how each company did, refer to the following plot.

```{r company-size-plot}
top_50 %>% 
  ggplot(aes(x = revenue_usd_billions, y = worldwide_box_office)) +
  geom_point(aes(color = company), alpha = .6, size = 4, ) +
  geom_smooth(se = F, method = "lm") +
  labs(
    title = "Company Revenue vs. Box Office Performance of Video Game Movies",
    x = "Yearly Company Revenue (Billions)",
    y = "Box Office Earnings (USD)",
    color = "Company"
  ) +
  scale_x_log10() +
  theme_minimal() +
  theme(legend.position = "bottom")
```

  Of the eight companies represented in this plot, we can see that the spread of movie success doesn't really change as x (the annual company revenue) increases. All of the companies seem to hover between the million and hundred-million range for box office performance with the exception of Nintendo's Super Mario and Super Mario Galaxy movies, which are special cases due to their significance. Because of these two outliers, the fitted line seems to show a general increase where one doesn't really exist.
  
  In summary, the size of the publishing company doesn't seem to play a noticable factor in the earnings of a movie based on one of their IPs. Larger companies do about the same as smaller ones.
  
## Question 3: Franchise Fatigue

[Reintroduce question]

[Insert Data Table]

[Explain Data Table]

[Insert Plot]

[Explain Plot]

[Concluding Statements]

## Conclusion

## References

[If needed]

## Code Appendix

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}


```

## AI Usage

**Harsh Rathi:**

> 

**Charles Bupp:**

> 

**Mason Mitchell:**

> 