---
title: "Video Game Film Adaptations"
author: "Harsh, Charlie, Mason"
date: "2026-06-15"
format: pdf
---

```{r setup}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(readr)
library(janitor)
library(scales)

ggplot2::theme_set(theme_minimal(base_size = 12))
```

## Research Questions

### Q1: Critic–Audience Agreement Over Time

Do critics and audiences agree on video game movies, and has that gap changed over time?

Compare `rotten_tomatoes` and `metacritic` against `cinema_score` (convert letter grades to numeric scale, drop `"N/A"` rows). Also plot the critic/audience gap across `release_date` to see if it narrows over time.

### Q2: Publisher Size & Box Office

Does publisher size predict box office success?

Join external publisher revenue/size data on `original_game_publisher`. Filter to USD only (`worldwide_box_office_currency == "$"`). Adjust `worldwide_box_office` for inflation to 2024 dollars via a CPI table joined on year extracted from `release_date`.

### Q3: Franchise Fatigue

Does franchise fatigue exist in video game movie series?

Use regex on `title` to detect both numbered sequels (`"Resident Evil 2"`) and subtitle-based sequels (`"Resident Evil: Apocalypse"`). Assign installment numbers within each franchise sorted by `release_date`. Focus on franchises with 3+ films. Plot inflation-adjusted USD box office and `rotten_tomatoes` by installment number.

### Known Data Issues

- Heavy NAs across score columns (`rotten_tomatoes`, `metacritic`, `cinema_score`)
- Mixed currencies in `worldwide_box_office` and `budget_low`/`budget_high`
- No inflation adjustment in raw data
- `cinema_score` is a character column needing numeric conversion before analysis

## Data Import

```{r import}
#| label: import

game_films <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-06-09/game_films.csv",
  show_col_types = FALSE
) |>
  janitor::clean_names()

glimpse(game_films)
```

## Data Preparation: Currency Standardization

All monetary columns are in their *original* currency. We standardize to USD.

```{r currency-lookup}
#| label: currency-lookup

fx_rates <- tibble::tribble(
  ~currency, ~to_usd,
  "$",        1,
  "¥",        0.0091,
  "£",        1.30,
  "HK$",      0.128,
  "NT$",      0.033
)
```

```{r currency-convert}
#| label: currency-convert

game_films <- game_films |>
  left_join(fx_rates, by = c("worldwide_box_office_currency" = "currency")) |>
  mutate(
    worldwide_box_office_usd = worldwide_box_office * to_usd
  ) |>
  select(-to_usd) |>
  left_join(fx_rates, by = c("budget_currency" = "currency")) |>
  mutate(
    budget_low_usd  = budget_low  * to_usd,
    budget_high_usd = budget_high * to_usd
  ) |>
  select(-to_usd)

summary(game_films$worldwide_box_office_usd)
```

## Data Preparation: Inflation Adjustment

Adjust all USD monetary columns to constant 2024 dollars using CPI-U.

```{r cpi-table}
#| label: cpi-table

cpi <- tibble::tribble(
  ~year, ~cpi,
  1986, 109.6,
  1987, 113.6,
  1988, 118.3,
  1989, 124.0,
  1990, 130.7,
  1991, 136.2,
  1992, 140.3,
  1993, 144.5,
  1994, 148.2,
  1995, 152.4,
  1996, 156.9,
  1997, 160.5,
  1998, 163.0,
  1999, 166.6,
  2000, 172.2,
  2001, 177.1,
  2002, 179.9,
  2003, 184.0,
  2004, 188.9,
  2005, 195.3,
  2006, 201.6,
  2007, 207.3,
  2008, 215.3,
  2009, 214.5,
  2010, 218.1,
  2011, 224.9,
  2012, 229.6,
  2013, 233.0,
  2014, 236.7,
  2015, 237.0,
  2016, 240.0,
  2017, 245.1,
  2018, 251.1,
  2019, 255.7,
  2020, 258.8,
  2021, 271.0,
  2022, 292.7,
  2023, 304.7,
  2024, 313.7
)

cpi_2024 <- cpi$cpi[cpi$year == 2024]
```

```{r inflation-adjust}
#| label: inflation-adjust

game_films <- game_films |>
  mutate(release_year = lubridate::year(release_date)) |>
  left_join(cpi, by = c("release_year" = "year")) |>
  mutate(
    inflation_factor = cpi_2024 / cpi,
    worldwide_box_office_usd_adj = worldwide_box_office_usd * inflation_factor,
    budget_low_usd_adj  = budget_low_usd  * inflation_factor,
    budget_high_usd_adj = budget_high_usd * inflation_factor
  ) |>
  select(-cpi, -inflation_factor)

game_films |>
  filter(!is.na(worldwide_box_office_usd)) |>
  summarise(
    n                    = n(),
    raw_median           = median(worldwide_box_office, na.rm = TRUE),
    usd_median           = median(worldwide_box_office_usd, na.rm = TRUE),
    adj_median           = median(worldwide_box_office_usd_adj, na.rm = TRUE)
  )
```

## Data Preparation: Movies Only!

Keep only entries that are theatrical releases. All others lack crucial information about budget, gross earnings, and review scores.

```{r theatrical-release}
#| label: theatrical-release

game_films <- game_films %>% 
  filter(category == "Theatrical release")

```

## Data Preparation: Cinemascore conversion

converts the Cinemascore variable to a numeric score

```{r cinema-convert}
#| label: cinema-convert

cinema_score_conversion <- tibble::tribble(
  ~cinemascore_letter, ~to_num,
  "N/A",        NA,
  "A+",        100,
  "A",        91.6666,
  "A-",      83.3333,
  "B+",      74.9999,
  "B",        66.6666,
  "B-",        58.3333,
  "C+",      49.9999,
  "C",      41.6666,
  "C-",        33.3333,
  "D+",      24.9999,
  "D",      16.6666,
  "D-",        8.3333,
  "F",        0,
)

game_films <- game_films %>% 
  left_join(cinema_score_conversion, join_by("cinema_score" == "cinemascore_letter"))

game_films <- game_films %>%
  mutate(cinema_score = to_num) %>% 
  select(-starts_with("to_num"))

```

## Cleaned Dataset Summary

```{r final-glimpse}
#| label: final-glimpse

glimpse(game_films)

game_films |>
  summarise(
    total_films        = n(),
    n_with_box_office  = sum(!is.na(worldwide_box_office)),
    n_with_rt          = sum(!is.na(rotten_tomatoes)),
    n_with_cinema      = sum(!is.na(cinema_score) & cinema_score != "N/A"),
    date_range          = paste(range(release_date, na.rm = TRUE), collapse = " to ")
  )
```
