---
title: "Video Game Film Adaptations — Factors in Box Office and Critical Reception"
author: "Harsh, Charlie, Mason"
date: "2026-06-15"
format: pdf
---

```{r setup}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(readr)
library(janitor)
library(scales)
library(knitr)
library(kableExtra)

ggplot2::theme_set(theme_minimal(base_size = 12))
```

## Introduction & Motivation

[Paragraphs explaining what video game film adaptations are, why they matter, the business/cultural context. Frame why analyzing critic-audience agreement, publisher size influence, and franchise fatigue is interesting and relevant.]

In this project, we aim to explore three core research questions:

1. Do critics and audiences agree on video game movies, and has that gap changed over time?
2. Does publisher size predict box office success?
3. Does franchise fatigue exist in video game movie series?

## Research Questions

### Q1: Critic–Audience Agreement Over Time

Do critics and audiences agree on video game movies, and has that gap changed over time?

Compare `rotten_tomatoes` and `metacritic` against `cinema_score` (convert letter grades to numeric scale, drop `"N/A"` rows). Also plot the critic/audience gap across `release_date` to see if it narrows over time.

### Q2: Publisher Size & Box Office

Does publisher size predict box office success?

Join external publisher revenue/size data on `original_game_publisher`. Filter to USD only (`worldwide_box_office_currency == "$"`). Adjust `worldwide_box_office` for inflation to 2024 dollars via a CPI table joined on year extracted from `release_date`.

### Q3: Franchise Fatigue

Does franchise fatigue exist in video game movie series?

Use regex on `title` to detect both numbered sequels (`"Resident Evil 2"`) and subtitle-based sequels (`"Resident Evil: Apocalypse"`). Assign installment numbers within each franchise sorted by `release_date`. Focus on franchises with 3+ films. Plot inflation-adjusted USD box office and `rotten_tomatoes` by installment number.

## Data Provenance

For this project, we used two data sources to explore video game film adaptations.

### Primary Dataset

- **Source**: TidyTuesday (GitHub: `rfordatascience/tidytuesday`)
- **Description**: The dataset was scraped from Wikipedia's "List of films based on video games." It contains 439 rows and 19 columns, with each row representing one film adaptation of a video game. Variables include title, release date, category (theatrical, direct-to-video, TV, etc.), original game publisher, worldwide box office, budget bounds, Rotten Tomatoes score, Metacritic score, and CinemaScore audience grade.
- **Purpose**: Originally compiled for the TidyTuesday weekly data project to enable exploration of video game film adaptations.
- **Cases**: Each row represents one film based on a video game.

```{r provenance-primary}
#| label: provenance-primary
#| echo: false

# Load and show structure
game_films <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-06-09/game_films.csv",
  show_col_types = FALSE
) |>
  janitor::clean_names()

# Key dimensions
cat("Primary dataset:", nrow(game_films), "films,", ncol(game_films), "variables\n")
```

### Supplementary Data

- **CPI Table**: U.S. Bureau of Labor Statistics CPI-U annual averages (series CUUR0000SA0), 1986–2024. Used to adjust box office figures to constant 2024 dollars.
- **Publisher Revenue Data**: [TBD — source to be determined; candidate: company financial filings, Wikipedia publisher revenue data, or FRED/industry databases]

## Data Wrangling & Cleaning

To address our research questions, we performed the following steps:

### Variable Selection

We retained variables directly relevant to our analysis: `title`, `release_date`, `category`, `original_game_publisher`, `worldwide_box_office`, `worldwide_box_office_currency`, `budget_low`, `budget_high`, `budget_currency`, `rotten_tomatoes`, `metacritic`, `cinema_score`, and `distributor`.

### Currency Standardization

All monetary columns were in their original currency. We standardized to USD using approximate exchange rates:

```{r wrangle-currency}
#| label: wrangle-currency
#| echo: false

fx_rates <- tibble::tribble(
  ~currency, ~to_usd,
  "$",        1,
  "¥",        0.0091,
  "£",        1.30,
  "HK$",      0.128,
  "NT$",      0.033
)
```

### Inflation Adjustment

Box office figures were adjusted to constant 2024 dollars using CPI-U. The factor for each film is `CPI_2024 / CPI_year`, where the year is extracted from `release_date`.

### Handling Missing Data

- `cinema_score`: rows with `"N/A"` are dropped for Q1 analysis.
- `rotten_tomatoes` / `metacritic`: rows with `NA` are excluded from relevant analyses.
- `worldwide_box_office`: analyses using box office exclude films with missing financial data.
- Currency conversion: films with non-USD currencies that lack matching exchange rates are flagged.
- `cinema_score` is a character column and must be converted to a numeric scale (A+ through F) before any quantitative analysis.

### Final Dataset

The cleaned dataset includes inflation-adjusted USD box office (`worldwide_box_office_usd_adj`), numeric cinema score mapping, and standardized currency columns, ready for analysis across all three research questions.

## Exploratory Data Analysis

To derive meaningful insights and address our research questions, we conducted extensive EDA. This included generating summary tables and creating visualizations to understand relationships between key variables and uncover patterns.

We started by constructing summary tables to provide an overview of key statistics. Next, we developed data visualizations:

- **Scatter plots**: To explore relationships between critic scores, audience scores, and box office.
- **Line plots**: To examine critic–audience gap over time.
- **Bar/box plots**: To compare publisher performance and franchise installment trends.
- **Correlation analysis**: To assess relationships among numeric variables.

Below, we detail the most critical tables and visualizations, along with the key insights they helped us derive.

---

### Descriptive Statistics

Table 1 provides an overview of the dataset's key numeric variables. The summary statistics (mean, median, minimum, maximum) offer context about the range and distribution of these variables.

**Table 1: Summary Statistics for Key Variables**

```{r eda-summary-table}
#| label: eda-summary-table
#| echo: false
#| eval: false

# PLACEHOLDER — replace with actual summary statistics
game_films |>
  select(worldwide_box_office_usd_adj, rotten_tomatoes, metacritic, cinema_score_numeric) |>
  summarise(across(everything(), list(
    mean   = ~mean(.x, na.rm = TRUE),
    median = ~median(.x, na.rm = TRUE),
    min    = ~min(.x, na.rm = TRUE),
    max    = ~max(.x, na.rm = TRUE)
  ))) |>
  pivot_longer(everything()) |>
  kable(caption = "Summary Statistics for Key Variables", digits = 2) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

Key Insights:
- [Discuss the range and distribution of box office values]
- [Note the spread of critic vs. audience scores]
- [Highlight the prevalence of NAs and what that implies]

---

### Q1: Critic–Audience Agreement

Figure 1 shows [describe the plot — e.g., scatter plot of rotten_tomatoes vs. cinema_score_numeric, or gap over time].

**Figure 1: [Caption describing plot — e.g., "Critic score vs. audience score for video game films, 1986–2024"]**

```{r eda-fig1}
#| label: eda-fig1
#| echo: false
#| eval: false

# PLACEHOLDER — Q1 plot
```

Key Insights:
- [Describe what the figure shows]
- [Discuss the critic–audience agreement/disagreement pattern]
- [Note whether the gap narrows over time]

---

Figure 2 examines [the critic–audience gap trend over time].

**Figure 2: [Caption — e.g., "Critic–audience score gap across release year"]**

```{r eda-fig2}
#| label: eda-fig2
#| echo: false
#| eval: false

# PLACEHOLDER — Q1 time series plot
```

Key Insights:
- [Discuss the trend over time]
- [Note any inflection points or eras]

---

### Q2: Publisher Size & Box Office

Table 2 compares box office performance across publishers, grouped by [publisher size / revenue tier].

**Table 2: [Caption — e.g., "Average inflation-adjusted box office by publisher size tier"]**

```{r eda-table2}
#| label: eda-table2
#| echo: false
#| eval: false

# PLACEHOLDER — publisher size comparison table
```

Key Insights:
- [Larger publishers tend to...]
- [Note any notable outliers or exceptions]

---

Figure 3 visualizes [publisher size vs. box office performance, possibly with 3+ variables using color/facets/size].

**Figure 3: [Caption — showing 3+ variables — e.g., "Box office vs. publisher revenue, colored by category and sized by critic score"]**

```{r eda-fig3}
#| label: eda-fig3
#| echo: false
#| eval: false

# PLACEHOLDER — multivariate figure (3+ variables)
```

Key Insights:
- [Discuss relationships among the multiple variables shown]
- [This figure satisfies the 3+ variables requirement]

---

### Q3: Franchise Fatigue

Figure 4 plots inflation-adjusted box office by installment number for franchises with 3+ films.

**Figure 4: [Caption — e.g., "Box office by installment number for major video game film franchises (inflation-adjusted 2024 USD)"]**

```{r eda-fig4}
#| label: eda-fig4
#| echo: false
#| eval: false

# PLACEHOLDER — franchise fatigue plot (box office)
```

Key Insights:
- [Does box office decline with later installments?]
- [Compare across franchises]

---

Figure 5 plots Rotten Tomatoes scores by installment number.

**Figure 5: [Caption — e.g., "Rotten Tomatoes score by installment number for major franchises"]**

```{r eda-fig5}
#| label: eda-fig5
#| echo: false
#| eval: false

# PLACEHOLDER — franchise fatigue plot (critic scores)
```

Key Insights:
- [Do critic scores decline with later installments?]
- [Is the pattern consistent across franchises?]

---

## Findings / Conclusions

This project explored three questions about video game film adaptations.

**Q1: Critic–audience agreement.** [Answer — do they agree? Has the gap narrowed over time?]

**Q2: Publisher size and box office.** [Answer — does publisher size predict financial success?]

**Q3: Franchise fatigue.** [Answer — does box office and/or critical reception decline with later installments?]

Overall, [broader takeaway about the video game film industry based on the analysis].

## Limitations & Future Work

- **Exchange rates**: Fixed approximate rates were used; historical daily rates would improve accuracy.
- **CPI data**: Annual averages were used; more precise adjustment by month could refine the inflation correction.
- **Missing data**: Many films lack scores or box office data, potentially biasing results toward larger, more prominent releases.
- **Publisher size**: The external revenue data source is [TBD] and may not perfectly reflect "franchise power."
- **Franchise detection**: Regex-based franchise extraction may misclassify ambiguous titles.
- **Future work**: Could incorporate genre data, streaming vs. theatrical splits, or audience demographics. A predictive model for box office using publisher and critic variables would extend the analysis.

## AI Use Statement

[Team member 1]: [1–2 sentences on how AI was used]
[Team member 2]: [1–2 sentences on how AI was used]
[Team member 3]: [1–2 sentences on how AI was used]

## References

- TidyTuesday (2026). Game Films dataset. https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-06-09
- U.S. Bureau of Labor Statistics. CPI-U (CUUR0000SA0), 1986–2024.
- [Publisher revenue source — TBD]
- [Any additional sources]

---

# Code Appendix

```{r codeAppend}
#| ref.label: knitr::all_labels()
#| echo: true
#| eval: false
```
